Local Type Inference for Polarised System F with Existentials

### 2023-10-30 oopsla24/tex/04-algorithm.tex
Finish the Algorithmic section

\begin{figure}[h]
  \hfill\\
  \begin{multicols}{2}
    \ottdefnANsub{}
    \columnbreak\\
    \ottdefnAPsup{}
  \end{multicols}
  \caption{Subtyping Algorithm}
  \label{fig:alg-subtyping}
\end{figure}

@@

\begin{figure}[h]
  \hfill
  \begin{multicols}{2}
  \ottdefnUNUnif{}
  \columnbreak\\
  \ottdefnUPUnif{}
  \end{multicols}
  \caption{Unification Algorithm}
  \label{fig:unification}
\end{figure}

@@

\begin{figure}[h]
  \ottdefnSCME\\
  \caption{Merge of Matching Constraint Entries}
  \label{fig:merge-entries}
\end{figure}

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnLUBUp{}
    \columnbreak\\
    \ottdefnLUBNsub{}
  \end{multicols}
  \caption{Type Upgrade and Leas Upper Bound Algorithms}
  \label{fig:type-upgrade}
\end{figure}

@@

\begin{figure}[h]
    \ottdefnAUAUP{}

    \vspace{\baselineskip}

    \ottdefnAUAUN{}
    \caption{Anti-Unification Algorithm}
    \label{fig:anti-unification}
\end{figure}

@@

        % - Overview (Dependency Graph)
        % *** The Inference Algorithm
        % *** Constraint Singularity

\begin{figure}
  \ottdefnATNInf{}
  \caption{Algorithmic Negative Type Inferences}
  \label{fig:type-inference}
\end{figure}

\begin{figure}
  \ottdefnATPInf{}
  \caption{Algorithmic Positive Type Inferences}
  \label{fig:type-inference-pos}
\end{figure}

\begin{figure}
  \ottdefnATSpinInf{}
  \caption{Algorithmic Application Type Inferences}
  \label{fig:type-inference-app}
\end{figure}

>>>>

\begin{figure}[h]
  \hfill\\
  \begin{multicols}{2}
    \ottdefnANsubLabeled{}
    \columnbreak\\
    \ottdefnAPsupLabeled{}
  \end{multicols}
  \caption{Subtyping Algorithm}
  \label{fig:alg-subtyping}
\end{figure}

@@

\begin{figure}[h]
  \hfill
  \begin{multicols}{2}
  \ottdefnUNUnifLabeled{}
  \columnbreak\\
  \ottdefnUPUnifLabeled{}
  \end{multicols}
  \caption{Unification Algorithm}
  \label{fig:unification}
\end{figure}

@@

\begin{figure}[h]
  \ottdefnSCMELabeled\\
  \caption{Merge of Matching Constraint Entries}
  \label{fig:merge-entries}
\end{figure}

@@

\begin{figure}[h]
  \begin{multicols}{2}
    \ottdefnLUBUpLabeled{}
    \columnbreak\\
    \ottdefnLUBNsubLabeled{}
  \end{multicols}
  \caption{Type Upgrade and Leas Upper Bound Algorithms}
  \label{fig:type-upgrade}
\end{figure}

@@

The anti-unification algorithm \cite{todo}, 
is a procedure dual to unification
For two given (potentially different) expressions, 
it finds the most specific generalizer---the most detailed pattern that 
matches both of the input expressions. As evidence, it can also return 
two substitutions that instantiate the `holes' of the pattern to the input expressions.

In our case, we have to be more demanding on the anti-unification algorithm.
Since we use it to construct an existential type, whose (negative) quantified variables
can only be instantiated with negative types, we must make sure that the pattern 
has `holes' only at negative positions. Moreover, we must make sure that
the resulting substitutions for the `holes' are well-formed in the initial contest, 
and do not contain variables bound later. For example, 
the anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and
$[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$
must result in a `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, 
with a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$.
But it cannot be more specific such as
$[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole cannot be positive)
or $[[∀β⁺.γ̂⁻]]$ 
(since the instantiation cannot capture the bound variable $[[β⁺]]$).

The algorithm that finds the most specific generalizer of two types
under required conditions is given in \cref{fig:anti-unification}.
It consists of two mutually recursive procedures:
the positive and the negative anti-unification. 
As the positive and the negative anti-unification procedures
are symmetric in their interface, let us discuss how to read
the positive judgment.

The positive anti-unification judgment has form
$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$.
As an input, it takes a context $[[Γ]]$, in which the `holes'
instantiations must be well-formed, 
and two positive types: $[[iP1]]$ and $[[iP2]]$;
it returns a tuple of four components:
$[[Ξ]]$---a set of `holes' represented by negative algorithmic variables,
$[[uQ]]$---a pattern represented as a positive algorithmic type,
whose algorithmic variables are in $[[Ξ]]$,
and two substitutions $[[aus1]]$ and $[[aus2]]$
instantiating the variables from $[[Ξ]]$ such that
$[[ [aus1]uQ = iP1 ]]$ and $[[ [aus2]uQ = iP2 ]]$.

\begin{figure}[h]
    \ottdefnAUAUPLabeled{}
    \hfill\\
    \hfill\\
    \ottdefnAUAUNLabeled{}
    \caption{Anti-Unification Algorithm}
    \label{fig:anti-unification}
\end{figure}

At the high level, the algorithm scheme follows
the standard approach \cite{todo} consisting of two principles:
\begin{enumerate}
  \item[(i)] if the input terms start with the same constructor,
    we anti-unify the corresponding parts recursively and 
    unite the results. This principle is followed by 
    all the rules except \ruleref{\ottdruleAUAULabel},
    which works as follows:
  \item[(ii)] if the first principle does not apply to the input terms $[[iN]]$
    and $[[iM]]$ (for instance, if they have different outer constructors), the
    anti-unification algorithm returns a `hole' such that one substitution maps
    it to $[[iN]]$ and the other maps it to $[[iM]]$. The name of this `hole'
    should have a name uniquely defined by the pair $([[iN]], [[iM]])$, so
    that it automatically merges with other `holes' mapped to the same
    pair of types, and thus, the initiality of the generalizer is ensured. 
\end{enumerate}

Let us discuss the specific rules of the algorithm in detail.

\paragraph{Variables} 
  Rules \ruleref{\ottdruleAUPVarLabel} and
  \ruleref{\ottdruleAUNVarLabel} generalize two equal variables.
  In this case, the resulting pattern is the variable itself,
  and no `holes' are needed.

\paragraph{Shifts} 
  Rules \ruleref{\ottdruleAUShiftDLabel} and
  \ruleref{\ottdruleAUShiftULabel} operate by congruence:
  they anti-unify the bodies of the shifts recursively and add
  the shift constructor back to the resulting pattern.

\paragraph{Quantifiers}
  Rules \ruleref{\ottdruleAUForallLabel} and
  \ruleref{\ottdruleAUExistsLabel} are symmetric. 
  They generalize two quantified types congruently, 
  similarly to the shift rules. 
  However, we also require that the quantified variables
  are fresh, and that the left-hand side variables are 
  equal to the corresponding variables on the right-hand side.
  To ensure it, alpha-renaming might be required in the
  implementation.

  Notice that the context $[[Γ]]$ is \emph{not} extended with 
  the quantified variables. In this algorithm, $[[Γ]]$ 
  does not play the role of a current typing context, but rather
  a snapshot of a context at the moment of calling the anti-unification,
  \ie the context in which the instantiations of the `holes' 
  must be well-formed.

\paragraph{Functions}
  Rule \ruleref{\ottdruleAUArrowLabel} congruently generalizes two function types. 
  An arrow type is the only binary constructor, 
  and thus, it is the only rule where the union of the anti-unification results is substantial.
  The interesting is the case when the resulting generalization of the 
  input types and the resulting generalization of the output types 
  have `holes' mapped to the same pair of types. 
  In this case, the algorithm must merge the `holes' into one.
  For example, the anti-unification of 
  $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$
  must result in $[[↓γ̂⁻ → γ̂⁻]]$,
  rather than $[[↓γ1̂⁻ → γ2̂⁻]]$.

  In our representation of the anti-unification algorithm, this `merge' happens
  automatically:
  following the rule \ruleref{\ottdruleAUAULabel},
  the name of the `hole' is uniquely defined by the pair of types it is mapped to.  
  Specifically, when anti-unifying $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$ our algorithm returns 
  $[[↓α̂⁻_{α⁻, β⁻} → α̂⁻_{α⁻, β⁻}]]$, that is a renaming of $[[↓γ̂⁻ → γ̂⁻]]$.

  This way, as the output the rule returns the following tuple:
  \begin{itemize}
    \item $[[Ξ1 ∪ Ξ2]]$---a simple union of the sets of `holes' 
      returned from by the recursive calls, 
    \item $[[uQ → uM]]$---the resulting pattern 
      constructed from the patterns returned recursively.
    \item $[[aus1 ∪ aus1']]$ and $[[aus2 ∪ aus'2]]$
      --- a union (in a relational sense)
      of the substitutions returned by the recursive calls. 
      It is worth noting that the union is well-defined because
      the result of the substation on a `hole' is determined by the 
      name of the `hole'.
  \end{itemize}

\paragraph{The Anti-Unification Rule}
  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
  algorithm. If the congruent rules are not applicable, 
  it means that the input types have a substantially different structure,
  and thus, the only option is to create a `hole'. 
  There are three important aspects of this rule that we would like to discuss.

  First, as mentioned earlier, the freshly created `hole' has a name
  that is uniquely defined by the pair of input types.
  It is ensured by the following invariant:
  all the `holes' in the algorithm have name $[[α̂⁻]]$ 
  indexed by the pair of negative types it is mapped to.
  This way, the returning set of `holes' is a singleton set
  $\{ [[ â⁻_{iN, iM} ]]\}$; the resulting 
  pattern is the `hole' $[[â⁻_{iN, iM}]]$,
  and the mappings simply send it to the corresponding types:
  $[[â⁻_{iN, iM} ↦ iN]]$ and $[[â⁻_{iN, iM} ↦ iM]]$.

  Second, this rule is only applicable to negative types, 
  moreover, the input types are checked to be well-formed 
  in the outer context $[[Γ]]$. This is required by the 
  usage of anti-unification: we 
  call it to build an existential type that would be 
  an upper bound of two input types via abstracting
  some of their subexpressions under existential quantifiers.
  The existentials quantify over \emph{negative} variables,
  and they must be instantiated in the context
  available at that moment.

  Third, the rule is only applicable if all other rules fail. 
  Notice that it could happen even when the input types have matching constructors. 
  For example, the generalizer of $[[↑α⁺]]$ and
  $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$),
  rather than $[[↑γ̂⁺]]$. This way, the algorithm must try to 
  apply the congruent rules first, and only if they fail,
  apply \ruleref{\ottdruleAUAULabel}.
  This principle makes the inference system not syntax-directed:
  it is not known a priori (before the recursive call) 
  whether the corresponding congruent rule or rule \ruleref{\ottdruleAUAULabel} 
  will be applied.

@@

Finally, we present the type inference algorithm. 
Similarly to the subtyping algorithm, it structurally corresponds to the declarative
inference specification, meaning that most of the algorithmic rules have
declarative counterparts, with respect to which they are sound and complete.

This way, the inference algorithm also consists of three mutually recursive procedures: 
the positive type inference, the negative type inference, and the
application type inference.

The positive and the negative type inference judgments have symmetric forms:
$[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ c : iN]]$. Both of these algorithms
take as an input typing context $[[Γ]]$, a variable context $[[Φ]]$, and 
a term (a value or a computation) taking its type variables from $[[Γ]]$, 
and term variables from $[[Φ]]$. As an output, they return a type
of the given term, which we guarantee to be normalized.

The application type inference judgment has form 
$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
and a constraint context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
a list of arguments (terms) $[[args]]$ the head is applied to.
The head may contain algorithmic variables specified by $[[Θ1]]$, 
in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
As a result, the application inference judgment returns 
$[[uM]]$---a normalized type of the result of the application.
Type $[[uM]]$ may contain new algorithmic variables, and thus, 
the judgment also returns $[[Θ2]]$---an updated constraint context
and $[[SC]]$---a set of subtyping constraints.
Together $[[Θ2]]$  and $[[SC]]$ specify how the algorithmic variables 
must be instantiated.

\begin{figure}
  \ottdefnATPInfLabeled{}
  \hfill \\
  \ottdefnATNInfLabeled{}
  \hfill \\
  \ottdefnATSpinInfLabeled{}
  \caption{Algorithmic Type Inference}
  \label{fig:type-inference}
\end{figure}

The inference rules are shown in
\cref{fig:type-inference}.
Next, we discuss them in detail.

\paragraph{Variables}
  Rule \ruleref{\ottdruleATVarLabel} 
  infers the type of a positive variable by looking it up in the 
  term variable context and normalizing the result.

\paragraph{Annotations}
  Rules \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
  are symmetric.
  First, they check that the annotated type is well-formed in the 
  given context $[[Γ]]$. Then they make a recursive call to infer the 
  type of annotated expression, check that the inferred type is a subtype of 
  the annotation, and return the normalized annotation.

\paragraph{Abstractions}
  Rule \ruleref{\ottdruleATtLamLabel} infers the type of a lambda abstraction.
  It checks the well-formedness of the annotation $[[iP]]$,
  makes a recursive call to infer the type of the body in the extended context, 
  and returns the corresponding arrow type.
  Since the annotation $[[iP]]$ is allowed to be non-normalized,
  the rule also normalizes the resulting type.

  Rule \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
  Similarly to the previous case, it makes a recursive call to infer the type
  of the body in the extended \emph{type} context. 
  After that, it returns the corresponding universal type. 
  It is also required to normalize the result.  
  For instance, if $[[α⁺]]$ does not occur in the body of the lambda,
  the corresponding $[[∀]]$ will be removed.

\paragraph{Return and Thunk}
  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
  are similar to the declarative rules: they make a recursive call
  to type the body of the thunk or the return expression and
  put the shift on top of the result.

\paragraph{Unpack}
  Rule \ruleref{\ottdruleATUnpackLabel}
  allows one to unpack an existential type.
  First, it infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
  and since the type is guaranteed to be normalized, binds 
  the quantified variables with $[[nas]]$.
  Then it infers the type of the body in the appropriately extended context
  and checks that the inferred type does not depend on $[[nas]]$
  by checking well-formedness $[[Γ ⊢ iN]]$.

\paragraph{Let Binders}
  Rule \ruleref{\ottdruleATVarLetLabel}
  represents the type inference of a standard let binder.
  It infers the type of the bound value $[[v]]$, and
  makes a recursive call to infer the type of the body in the extended context.

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  infers a type of \emph{annotated} applicative let binder.
  First, it infers the type of the head of the application,
  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
  After that, it makes a recursive call
  to the application inference procedure,
  returning an algorithmic type $[[↑uQ]]$, 
  that must be instantiated to a subtype of 
  the annotation $[[↑uP]]$.
  Then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
  check whether the instantiation to the annotated type $[[↑iP]]$ is possible,
  and if it is, the algorithm infers the type of the body in the extended context,
  and returns it as the result.

  Rule \ruleref{\ottdruleATAppLetAnnLabel}
  works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
  However, since there is no annotation, 
  instead of checking
  that the inferred type $[[↑uQ]]$ 
  can be a subtype of the annotation,
  the algorithm checks that $[[↑uQ]]$ is unique. 
  As we prove, uniqueness means that all the algorithmic variables of 
  $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$.
  This way, uniqueness is checked by the combination of
  $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
  Together, these two premises guarantee that the only 
  possible instantiation of $[[uQ]]$ is $[[ [uσ]uQ ]]$.

\paragraph{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of applied arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  Rule \ruleref{\ottdruleATForallAppLabel},
  analogously to the declarative case,
  is the rule ensuring the implicit elimination of the universal quantifiers. 
  This is the place where the algorithmic variables are introduced.
  The algorithm simply replaces the quantified variables 
  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
  and makes a recursive call in the extended context.

  To ensure that this step does not cause infinite recursion, 
  we also check that the head type has at least one 
  $[[∀]]$-quantifier. Also, to force the algorithm to
  apply rule \ruleref{\ottdruleATEmptyAppLabel} when  
  there are no arguments, we require $[[args ≠ ·]]$.

\paragraph{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the minimal constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.
